---
title: "Lab 6 - Alternative"
subtitle: "Contributing to R for Data Science"
editor: source
---

The exercises from the functions below were pulled from the newest version of
[*R for Data Science*](https://r4ds.hadley.nz/). Specifically, from Chapters 25 
and 26. For this "alternative" lab you will complete the exercises from the 
textbook, with the option of submitting a pull request to the 
[repository for the textbook solutions](https://github.com/mine-cetinkaya-rundel/r4ds-solutions/).

```{r}
#| message: false
library(tidyverse)
library(nycflights13)
```


# Vector Functions

**Question 1:** The `rescale01()` function below performs a min-max scaling to 
standardize a numeric vector, but infinite values are left unchanged. Rewrite
`rescale01()` so that `-Inf` is mapped to 0, and `Inf` is mapped to 1?
*Hint: This seems like a great place for `case_when()`!

```{r}
#| label: question-1

rescale01 <- function(x) {
  if (!is.vector(x)) {
    stop("The input for `x` must be a vector.")
  }

  
  x <- case_when(
    x == Inf ~ 1,
    x == -Inf ~ 0,
    TRUE ~ x
  )
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

rescale01(c(1, 2, 3, 4, 5, Inf, -Inf))
```


**Question 2:** Write a function that accepts a vector of birthdates and 
computes the age of each person in years.

```{r}
#| label: question-2
age_calc <- function(birthdates) {
  if (!is.vector(birthdates)) {
    stop("The input for `birthdates` must be a vector.")
  }
  
  birthdates <- as.Date(birthdates)
  today <- Sys.Date()
  ages <- as.integer((today - birthdates) / 365.25) 
  return(ages)
}

age_calc(c("2001-08-30", "2003-08-08", "2005-08-31", "1100-03-07"))
```

**Question 3:** Write a function that computes the variance and skewness of a
numeric vector. Feel free to look up the definitions on Wikipedia or elsewhere!

```{r}
#| label: question-3
var_skew <- function(x){
  if (!is.vector(x)) {
    stop("The input for `x` must be a vector.")
  }
  
  variance <- var(x)
  
  n <- length(x)
  mean <- mean(x)
  sd <- sd(x)
  
  skewness <- (sum((x - mean)^3)/(n * sd^3))
  return(list(variance = variance, skewness = skewness))
}

var_skew(c(1,2,3,4,11037))
```

**Question 4:** Write a function called `both_na()` which takes two vectors of
the same length and returns the number of positions that have an `NA` in both
vectors.

```{r}
#| label: question-4
both_na <- function(V1, V2){
  if (!is.vector(V1) | !is.vector(V2)) {
    stop("The input for `V1` and 'V2' must be a vector.")
  }
  
  if (length(V1) != length(V2)) {
    stop("make those vectors the same length or ELSE")
  }
  
  total_both_na_positions <- sum(is.na(V1) & is.na(V2))
  return(total_both_na_positions)
  
}

both_na(c(NA, NA, 11037), c(NA, 1, 2))
```


## Data Frame Functions

**Question 5:** Insert the data frame function you wrote from Lab 6 (either
Exercise 1 or Exercise 2). 

```{r}
#| label: function-from-lab-2

remove_outliers <- function(df, ..., sd_thresh = 3){
  
  #error cases for df and sd_thresh
  if (!is.data.frame(df)) {
    stop("df must be a data frame please!!!!!!!")
  }
  if (!is.numeric(sd_thresh) | sd_thresh <= 0) {
    stop("sd_thresh has to be a positive number please!!!!!!")
  }
  
  #turn everything in ellipses to quosures
  names_quosures <- enquos(...)
  #take everything in names_quosures and extracts the names as character strings
  names <- map_chr(names_quosures, quo_name)
  
  #error cases for the names/...
  if (length(names) < 1) {
    stop("provide column names to remove outliers from please bro")
  }
  if (!is.character(names)) {
    stop("column names must be character strings!! ")
  }
  if (!all(names %in% names(df))) {
    stop("the column(s) are not in the data frame :(")
  }
  
  #meat of the function. check if the col is numeric, calc z score and then . way to do this without for looping?
  df_RO <- df
  for (name in names) {
    if (!is.numeric(df_RO[[name]])) {
      warning(paste("Skipping non-numeric column ", name))
      next
    }
    mean <- mean(df_RO[[name]], na.rm = TRUE)
    sd <- sd(df_RO[[name]], na.rm = TRUE)
    abs_z <- abs((df_RO[[name]] - mean) / sd)
    df_RO <- df_RO[abs_z <= sd_thresh | is.na(abs_z), ]
  }
  return(df_RO)
}

## Testing how your function handles multiple input variables
remove_outliers(diamonds, 
                price, 
                x, 
                y, 
                z)

## Testing how your function handles an input that isn't numeric
remove_outliers(diamonds, 
                price, 
                color)

## Testing how your function handles a non-default sd_thresh
remove_outliers(diamonds, 
                price,
                x, 
                y, 
                z, 
                sd_thresh = 2)
```

For Questions 6 - 10 you will write different functions which work with data 
similar to the `nycflights13` data. 

```{r}
#| include: false
df <- head(nycflights13::flights)
```


**Question 6:** Write a `filter_severe()` function that finds all flights that
were cancelled (i.e. `is.na(arr_time)`) or delayed by more than an hour.

```{r}
#| label: question-6

#assuming arrival or departure delay
filter_severe <- function(df, arrival_delay_var, departure_delay_var, arrival_time_var){

#checks to make sure all the supplied columns are numeric
    non_num <- df %>%
    select( {{ arrival_delay_var }}, {{departure_delay_var}}, {{arrival_time_var}}) %>%
    sapply(is.numeric)

  if (!all(non_num)) {
    stop(
      str_c("One of the columns you entered is not numeric.")
      )
  }
  
  
  filtered_df <- df %>%
    filter(is.na({{arrival_time_var}}) | {{arrival_delay_var}} > 60 | {{departure_delay_var}} > 60)
  
  return(filtered_df)
}

filter_severe(flights, arr_delay, dep_delay, arr_time)
```



**Question 7:** Write a `summarize_severe()` function that counts the number of 
cancelled flights and the number of flights delayed by more than an hour.

```{r}
#| label: question-7
summarize_severe <- function(df, arrival_delay_var, departure_delay_var, arrival_time_var){
  
  
#checks to make sure all the supplied columns are numeric
    non_num <- df %>%
    select( {{ arrival_delay_var }}, {{departure_delay_var}}, {{arrival_time_var}}) %>%
    sapply(is.numeric)

  if (!all(non_num)) {
    stop(
      str_c("One of the columns you entered is not numeric.")
      )
  }
  
  
  cancelled_count <- df %>%
    filter(is.na({{arrival_time_var}})) %>%
    count()
  
  delayed_count <- df %>%
    filter({{arrival_delay_var}} > 60 | {{departure_delay_var}} > 60) %>%
    count()
  
  output <- paste0("There are ", cancelled_count, " cancelled flights and there are ", delayed_count, " delayed flights, counting both arrival and departure delays.")
  return(output)
  
}

summarize_severe(flights, arr_delay, dep_delay, arr_time)
```

**Question 8:** Modify your `filter_severe()` function to allow the user to 
supply the number of hours that should be used to filter the flights that were
cancelled or delayed. 

```{r}
#| label: question-8

#assuming arrival or departure delay
filter_severe <- function(df, arrival_delay_var, departure_delay_var, arrival_time_var, hours = 1){

#checks to make sure all the supplied columns are numeric
    non_num <- df %>%
    select( {{ arrival_delay_var }}, {{departure_delay_var}}, {{arrival_time_var}}) %>%
    sapply(is.numeric)

  if (!all(non_num)) {
    stop(
      str_c("One of the columns you entered is not numeric.")
      )
  }
    
  if (!is.numeric({{hours}})) {
    stop(
      str_c("You must use a number for the `hours` input!")
      )
  }
  
  
  filtered_df <- df %>%
    filter(is.na({{arrival_time_var}}) | {{arrival_delay_var}} > hours*60 | {{departure_delay_var}} > hours*60)
  
  return(filtered_df)
}

filter_severe(flights, arr_delay, dep_delay, arr_time, hours=3)
```

**Question 9:** Write a `summarize_weather()` function that summarizes the
weather to compute the minimum, mean, and maximum, of a user supplied variable. 

```{r}
#| include: false
df <- head(nycflights13::weather, n=100)

nrow(nycflights13::weather)
```


```{r}
#| label: question-9

summarize_weather <- function(df, variable){
  
  non_num <- df %>%
    pull( {{ variable }}) %>%
    is.numeric()
  
   if (!non_num) {
    stop(
      str_c("You must use a numeric variable for the `variable` input!")
      )
  }
  
  df %>%
    summarize(
      min = min({{variable}}, na.rm = TRUE),
      mean = mean({{variable}}, na.rm = TRUE),
      max = max({{variable}}, na.rm=TRUE)
    )
    
}

summarize_weather(weather, temp)
```



**Question 10:** Write a `standardize_time()` function that converts the user
supplied variable that uses clock time (e.g., `dep_time`, `arr_time`, etc.) into
a decimal time (i.e. hours + (minutes / 60)).

```{r}
#| include: false
df <- head(nycflights13::flights, n=1000)
```

```{r}
#| label: question-10
standardize_time <- function(df, clock_time){
  non_num <- df %>%
    pull({{ clock_time }}) %>%
    is.numeric()
  
   if (!non_num) {
    stop(
      str_c("You must use a numeric variable for the `clock_time` input!")
      )
  }
  
  df %>%
    mutate(
      {{clock_time}} := case_when(
        !is.na({{clock_time}}) ~ floor({{clock_time}}/100) + ({{clock_time}} %% 100) / 60
      )
    )
  
  
}

standardize_time(flights, dep_time)

```


# Plotting Functions

You might want to read over the [Plot Functions section of *R for Data Science*](https://r4ds.hadley.nz/functions.html#plot-functions)

**Question 11:** Build a `sorted_bars()` function which:

- takes a data frame and a variable as inputs and returns a **vertical** bar
chart 
- sorts the bars in decreasing order (largest to smallest)
- adds a title that includes the context of the variable being plotted

*Hint 1: The `fct_infreq()` and `fct_rev()` functions from the forcats package will be helpful for sorting the bars!*
*Hint 2: The `englue()` function from the rlang package will be helpful for adding a variable's name into the plot title!*

I also made it so that if a variable had more than 10 unique values, the bar chart would be horizontal since the names of each category tend to blend together when the bar chart is vertical.

```{r}
#| label: question-11

sorted_bars <- function(df, var){
  #check if var is correct data type
  non_char_ord <- df %>%
    pull({{ var }}) %>%
    (\(x)is.character(x) | is.ordered(x) | is.factor(x))
  
   if (!non_char_ord) {
    stop(
      str_c("You must use a character or factor for the `var` input!")
      )
  }
  #use enquo to pull name of var
  var_name <- rlang::as_name(enquo(var))
  
  #count unique # of columns to determine if coord flipping later
  total_vals <- n_distinct(df %>% pull({{ var }}))
  
  p <- df %>%
    ggplot(aes(x = fct_infreq(as.factor({{var}})))) +
    geom_bar() +
    labs(
      title = rlang::englue("Distribution of { var_name }"),
      x = var_name,
      y = "Count"
    ) +
    theme_minimal() +
    theme(
      panel.grid.major.x = element_blank()
    )
  
  if (total_vals > 10) {
    p <- p + coord_flip()
  }
  
  return(p)
}

sorted_bars(diamonds, clarity)
```

# Iteration

Alright, now let's take our plotting function and iterate it! 

**Question 12:** Make a sorted barplot for every character variable in the `mpg`
dataset (built into `ggplot2`). 

```{r}
#| label: question-12

char_vars <- mpg %>%
  select(where(is.character)) %>%
  names()

map(char_vars, ~sorted_bars(mpg, !!sym(.x)))
```

# Contributing to the R for Data Science Community!

The functions you wrote for exercises 1-10 came from *R for Data Science*. You
could consider making a pull request to the repository for the solutions! 

<https://github.com/mine-cetinkaya-rundel/r4ds-solutions>

To learn more about how to make a pull request I would suggest this article: <https://usethis.r-lib.org/articles/pr-functions.html>
